---
title: "TagSeq_reanalysis"
author: "Kara Ryan"
date: "2025-05-27"
output:
  pdf_document: default
  html_document: default
---

```{r libs, echo=F}
library(dplyr)
library(tidyr)
library(stringr)
library(DESeq2)
library(ggplot2)
library(VennDiagram)
library(topGO)
library(gridExtra)

#BiocManager::install("clusterProfiler")
#BiocManager::install("pathview")
#BiocManager::install("enrichplot")
library(clusterProfiler)
```

# Analysis of TagSeq data after injection with AP39

### Experimental design:

Injection treatments:\
A -- control\
B -- PBS injection control\
D -- high AP39

All factors in the analysis:\
2 tissues\
2 river drainages\
2 pops per drainage\
3 treatments per pop

Separate for each tissue (L&G) and for each population (PuyS, PuyNS, TacS, TacNS) we will need:\
Lists of DEG for A vs B (control vs PBS)\
Lists of DEG for (A+B)/2 vs D (Control + PBS) vs AP39 injection\

### Read in count and annotation data

```{r input data}
# Read in count data and annotation
# counts <- read.table("/Users/kararyan/Documents/AP39_Tagseq/downsampled_count_M_cleaned_2_trans_removed.txt", skip = 1, header= TRUE)

counts <- read.table("/Users/kararyan/Downloads/all_samples_featurecounts.reheader.mx", header= TRUE)


#counts <- read.table("/Users/kararyan/Documents/AP39_Tagseq/Tagseq_counts_1_unstranded_ID.txt", header= TRUE)

anno <- read.csv("/Users/kararyan/Documents/AP39_Tagseq/metadata3.csv", stringsAsFactors = T)


anno<- as.data.frame(anno)
anno$pop <- as.factor(anno$pop)
anno$treatment <- as.factor(anno$treatment)
anno$tissue <- as.factor(anno$tissue)
anno %>%
  select(tissue, pop, treatment, group) %>%
  group_by(pop) %>%
  group_split() %>%
  lapply(summary)

summary(anno$group)
```


```{r input data}
# Make sample name the column name in the counts matrix 
colnames(counts)<-gsub("X.","",colnames(counts))

# Remove tissue swap samples
counts %>%
  select(-EFR22.080, -EFR22.081) -> counts

# View count data
ncol(counts)
nrow(counts)
tail(counts)
head(counts)

# Read in annotation data 
Pmex_human <- read.csv('/Users/kararyan/Documents/AP39_Tagseq/Table_S2_Poecilia_mexicana_Annotations.csv', header = T)

# read in human annotations 
Pmex_human %>%
  select(gene.ID, gene.name, Subject.sequence.ID, Protein.annotations) %>%
  separate(col = Subject.sequence.ID, into = c("sp", "Uniprot", "Hugo_human"), sep = '\\|') %>%
  select(-sp) %>%
  mutate(Hugo_human = str_remove_all(Hugo_human, "_HUMAN")) -> Pmex_human
anno
anno %>%
  filter(tube.id == "EFR22.108")
```

## Tissue split

```{r tissue split}

# filter metadata to gill tissue only 
anno %>%
  filter(tube.id !='EFR22.081'& tube.id != 'EFR22.080' & treatment != 'AP39.half' ) -> anno_filtered

# create a list of tube.ids to keep 
keep <- anno_filtered$tube.id
keep
#filter counts matrix to keep samples in gill list
rownames(counts) <- counts$Geneid
counts %>%
  select(all_of(keep)) -> counts_2
counts_2 <- counts_2[rowSums(counts_2 > 5) > 7, ]

# get sample size of unfiltered data
ncol(counts_2)

# convert gill counts to a matrix
counts_matrix <- as.matrix(counts_2)

# remove tube id column and convert it to the rownames 
anno_matrix <- as.matrix(anno_filtered[,-1])
rownames(anno_matrix) <- anno_filtered[,1]

# confirm the same number of samples in both the counts and annotation
all(rownames(anno_matrix %in% colnames(counts_matrix)))
# confirm the sample names are in the matrix 
all(rownames(anno_matrix) == colnames(counts_matrix))

anno

dds <- DESeqDataSetFromMatrix(countData = counts_matrix, colData = anno_matrix, 
                              design = ~ pop + treatment + tissue)

dds_all <- DESeq(dds)

# Use variance stabilizing transformation (VST) or rlog (vst is faster for large datasets)
vsd <- vst(dds_all, blind = TRUE)

# Create PCA plot using DESeq2's plotPCA function
pca_data <- plotPCA(vsd, intgroup = c("tissue", "pop"), returnData = TRUE, ntop = 2500)  # change "condition" as needed

# Compute % variance explained by each PC
percentVar <- round(100 * attr(pca_data, "percentVar"))
anno
pca_data %>%
  left_join(anno, by = c("name" = "tube.id"))%>%
  na.omit()->pca_data 

pca_data

# Plot with ggplot2
fig_s1<- ggplot(pca_data, aes(PC1, PC2, color = tissue.x, shape = pop.x)) +
  geom_point()+
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  scale_shape_manual(values = c(0,17, 15,2)) +
  theme_minimal() +
  labs(title = "PCA of VST-Transformed Counts (n=2,500 genes)") + 
  guides(color=guide_legend(title="Tissue")) + 
  guides(shape=guide_legend(title="Population"))

pca_data %>%
  arrange(PC1)

#pdf("/Users/kararyan/Documents/AP39_Tagseq/FigS1.pdf")
fig_s1
```


```{r}
dds <- estimateSizeFactors(dds)
normalized_counts <- counts(dds, normalized=TRUE)

# make the results a data frame
normalized_counts  <- as.data.frame(normalized_counts)
# make a column of the gene names 
normalized_counts$gene <- rownames(normalized_counts)
# join with annotations 
normalized_counts  %>% 
  left_join(Pmex_human,join_by('gene' =='gene.ID')) ->normalized_counts

row.names(normalized_counts)

normalized_counts %>%
  filter(gene.name == "sqrdl" | gene.name == 'LOC106917689' | gene.name == 'LOC106917690') -> sqr_pathway_counts
sqr_pathway_counts

sqr_pathway_counts %>%
  pivot_longer(!c(Protein.annotations, Hugo_human, Uniprot, gene.name, gene), names_to = "tube.id", values_to = "count") -> sqr_pathway_counts
sqr_pathway_counts

sqr_pathway_counts %>%
  left_join(anno) -> sqr_pathway_counts
sqr_pathway_counts %>%
  select(!c(treatment.c, treatment.c2, habitat2, drainage2, tissue2)) -> sqr_pathway_counts
sqr_pathway_counts



```


```{r}
sqr_pathway_counts %>%
         filter(gene.name == 'sqrdl') -> sqrdl

sqrdl$pop.treatment <- paste(sqrdl$pop, sqrdl$treatment, sep = ".")
color_palette <- c("s" = "gold", "ns" = "skyblue")

sqrdl$fill_color <- ifelse(sqrdl$treatment == "AP39.full",
                                  sqrdl$habitat,
                                  NA)

sqr_plot<-ggplot(sqrdl, aes(x = pop.treatment, y = count, color = habitat)) +
  geom_boxplot() +
  scale_color_manual(values = color_palette) +
  facet_wrap(~tissue) +
  theme_minimal()+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("SQR normalized counts")

sqr_pathway_counts %>%
         filter(gene.name == 'LOC106917689') -> ethe1_89

ethe1_89$pop.treatment <- paste(ethe1_89$pop, ethe1_89$treatment, sep = ".")
color_palette <- c("s" = "gold", "ns" = "skyblue")

ethe1_89$fill_color <- ifelse(ethe1_89$treatment == "AP39.full",
                                  ethe1_89$habitat,
                                  NA)

ethe1_89_plot<-ggplot(ethe1_89, aes(x = pop.treatment, y = count, color = habitat)) +
  geom_boxplot() +
  scale_color_manual(values = color_palette) +
  facet_wrap(~tissue) +
  theme_minimal()+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("ethe1 1 normalized counts")
ethe1_89_plot

sqr_pathway_counts %>%
         filter(gene.name == 'LOC106917690') -> ethe1_90

ethe1_90$pop.treatment <- paste(ethe1_90$pop, ethe1_90$treatment, sep = ".")
color_palette <- c("s" = "gold", "ns" = "skyblue")

ethe1_90$fill_color <- ifelse(ethe1_90$treatment == "AP39.full",
                                  ethe1_90$habitat,
                                  NA)

ethe1_90_plot<-ggplot(ethe1_90, aes(x = pop.treatment, y = count, color = habitat)) +
  geom_boxplot() +
  scale_color_manual(values = color_palette) +
  facet_wrap(~tissue) +
  theme_minimal()+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("ethe1 2 normalized counts")
ethe1_90_plot

sqr_pathway_counts$pop.treatment <- paste(sqr_pathway_counts$pop, sqr_pathway_counts$treatment, sep = ".")

sqr_pathway_counts$pop.treatment <- factor(sqr_pathway_counts$pop.treatment, levels = c('bonita.control', 'bonita.PBS', 'bonita.AP39.full', "pso.control", "pso.PBS", "pso.AP39.full", "puya.control", 'puya.PBS', "puya.AP39.full", 'lluvia.control', 'lluvia.PBS', 'lluvia.AP39.full'))

sqr_pathway_counts_plot <-ggplot(sqr_pathway_counts, aes(x = pop.treatment, y = count, color = habitat)) +
  geom_boxplot() +
  scale_color_manual(values = color_palette) +
  facet_wrap(~gene.name + tissue, ncol = 2) +
  theme_minimal()+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("SQR pathway genes normalized gene expression")
sqr_pathway_counts_plot
pdf("/Users/kararyan/Documents/AP39_Tagseq/sqr_pathway_genes.pdf")
sqr_pathway_counts_plot
sqr_pathway_counts %>%
  filter(pop.treatment == "bonita.AP39.full" & tissue == 'liver' & gene.name == "LOC106917690")
sqr_pathway_counts_plot_free_y <-ggplot(sqr_pathway_counts, aes(x = pop.treatment, y = count, color = habitat)) +
  geom_boxplot() +
  scale_color_manual(values = color_palette) +
  facet_wrap(~gene.name + tissue, ncol = 2, scales = 'free_y') +
  theme_minimal()+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("SQR pathway genes normalized gene expression")
sqr_pathway_counts_plot_free_y
pdf("/Users/kararyan/Documents/AP39_Tagseq/sqr_pathway_genes_free_y.pdf")
sqr_pathway_counts_plot_free_y

library(ggpubr)
ggarrange(sqr_plot + rremove("xlab"),  ethe1_89_plot + rremove("xlab"), ethe1_90_plot, ncol = 1, common.legend = FALSE )


ecotype_boxplot <-ggplot(sqr_pathway_counts, aes(x = treatment, y = count, color = habitat)) +
  geom_boxplot() +
  scale_color_manual(values = color_palette) +
  facet_wrap(~gene.name + tissue, ncol = 2, scales = 'free_y') +
  theme_minimal()+ theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("SQR pathway genes normalized gene expression")
ecotype_boxplot
```


```{r}
# Force fill_color to character (important!)
sqrdl$fill_color <- ifelse(sqrdl$treatment == "AP39.full",
                           as.character(sqrdl$habitat),
                           NA_character_)

# Define color palette for "s" and "ns"
color_palette <- c("s" = "gold", "ns" = "skyblue")

# Plot
ggplot(sqrdl, aes(x = pop.treatment, y = count, color = habitat, fill = fill_color)) +
  geom_boxplot() +
  scale_color_manual(values = color_palette) +
  scale_fill_manual(values = c(color_palette)) +  # you can remove "none" if not using it
  facet_wrap(~tissue) +
  theme_minimal()
```



## heatmap tissues 
```{r}

dds_all <- estimateSizeFactors(dds_all)
# Variance Stabilizing Transformation
vsd <- vst(dds_all, blind = TRUE)

# Extract the transformed expression matrix
vsd_mat <- assay(vsd)

# Calculate the row variance for each gene
gene_variances <- apply(vsd_mat, 1, var)

# Select top 2500 most variable genes
top_genes <- head(order(gene_variances, decreasing = TRUE), 2000)
top_vsd_mat <- vsd_mat[top_genes, ]

# Optional: Add annotation for columns (samples)
# Here we use the colData from the DESeq object
sample_anno <- as.data.frame(colData(dds))

library(pheatmap)
# Generate the heatmap with hierarchical clustering

sample_anno
sample_anno %>%
  select(tissue) -> sample_anno_2

pdf("heatmap_top2000.pdf", width = 10, height = 12)
pheatmap(top_vsd_mat,
         scale = "row",                 # Standardize each gene (z-score)
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "complete",
         annotation_col = sample_anno_2, # Sample metadata
         show_rownames = FALSE,        # Hide gene names for clarity
         fontsize_col = 10,            # Sample name font size
         main = "Top 2000 Variable Genes Heatmap")

# Save as PDF if desired

# pheatmap(...)  # Repeat the call inside
dev.off()
```














# Gill Analysis

### Set up gill matrix

```{r gill set up deseq object}

# filter metadata to gill tissue only and remove sample swap
anno %>%
  filter(tissue =='gill' & tube.id != 'EFR22.080' & treatment != "AP39.half" & tube.id != "EFR22.142") %>%
  select(tube.id, pop, treatment, group, mass, drainage, habitat)-> gill_meta
gill_meta

# create a list of tube.ids to keep 
keep <- gill_meta$tube.id

#filter counts matrix to keep samples in gill list
rownames(counts) <- counts$Geneid
counts

counts %>%
  select(all_of(keep)) -> gill_counts
gill_counts <- gill_counts[rowSums(gill_counts > 3) > 3, ]

# get sample size of unfiltered gill data
ncol(gill_counts)

nrow(gill_counts)
# convert gill counts to a matrix
gill_counts_matrix <- as.matrix(gill_counts)

# remove tube id column and convert it to the rownames 
gill_anno_matrix <- as.matrix(gill_meta[,-1])
rownames(gill_anno_matrix) <- gill_meta[,1]

# confirm the same number of samples in both the counts and annotation
all(rownames(gill_anno_matrix %in% colnames(gill_counts_matrix)))
# confirm the sample names are in the matrix 
all(rownames(gill_anno_matrix) == colnames(gill_counts_matrix))

ncol(gill_counts_matrix) 
nrow(gill_anno_matrix)

gill_meta$pop <- as.factor(gill_meta$pop)
summary(gill_meta$group)
```

### Set up DESeq design

```{r deseq gill}
summary(anno$mass)

# set up deseq2 dataset 
dds_gill <- DESeqDataSetFromMatrix(countData = gill_counts_matrix, colData = gill_anno_matrix, design = ~ treatment + pop + mass )
# Run DESeq2 
dds_gill <- DESeq(dds_gill)

vsd_gill <- vst(dds_gill, blind=T)

pcaData_gill<- plotPCA(vsd_gill, intgroup=c("treatment", 'pop'), returnData=TRUE, ntop = 10000) 
percentVar <- round(100 * attr(pcaData_gill, "percentVar")) 
pcaData_gill$fill_color <- ifelse(pcaData_gill$treatment == "AP39.full",
                                  pcaData_gill$habitat,
                                  NA)
# Step 2: Define color and fill palettes
color_palette <- c("s" = "gold", "ns" = "skyblue")  # change to actual habitat names
shape_palette <- c("taco" = 22, "puya" = 24)
# Use shape values 21 and 24 (can be filled)

gill_PCA <- ggplot(pcaData_gill, aes(x = PC1, y = PC2,
                                     color = habitat,
                                     shape = drainage,
                                     fill = fill_color)) +
  geom_point(size = 3, stroke = 1) +
 # geom_text(aes(label = name)) +
  scale_shape_manual(values = shape_palette) +
  scale_color_manual(values = color_palette) +
  scale_fill_manual(values = color_palette, na.value = NA) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme_minimal() +
  ggtitle("Gill PCA, n = 200") 

gill_PCA 
```

```{r}

dds_gill <- estimateSizeFactors(dds_gill)
# Variance Stabilizing Transformation
vsd_gill <- vst(dds_gill, blind = TRUE)

# Extract the transformed expression matrix
vsd_mat_gill <- assay(vsd_gill)

# Calculate the row variance for each gene
gene_variances_gill <- apply(vsd_mat_gill, 1, var)

# Select top 2500 most variable genes
top_genes_gill <- head(order(gene_variances_gill, decreasing = TRUE), 100)
top_vsd_mat_gill <- vsd_mat_gill[top_genes_gill, ]

# Optional: Add annotation for columns (samples)
# Here we use the colData from the DESeq object
sample_anno_gill <- as.data.frame(colData(dds_gill))

sample_anno_gill
# Generate the heatmap with hierarchical clustering

sample_anno_gill %>%
  select(treatment, habitat, drainage) -> sample_anno_2_gill

#pdf("heatmap_top2000.pdf", width = 10, height = 12)
pheatmap(top_vsd_mat_gill,
         scale = "row",                 # Standardize each gene (z-score)
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "complete",
         annotation_col = sample_anno_2_gill, # Sample metadata
         show_rownames = FALSE,        # Hide gene names for clarity
         fontsize_col = 10,            # Sample name font size
         main = "Top 2000 Variable Genes Heatmap")

# Save as PDF if desired

# pheatmap(...)  # Repeat the call inside
#dev.off()
```








```{r}
# Extract Cook's distance
cooks <- assays(dds_gill)[["cooks"]]

cooks <- as.data.frame(cooks)
cooks$gene <- row.names(cooks)
summary(cooks)
# Convert to tidy format for summarization
cooks_tidy <- cooks %>%
  pivot_longer(-gene, names_to = "sample", values_to = "cooks_distance")
```


```{r}
cooks_tidy
# Merge group information
cooks_tidy <- cooks_tidy %>%
  left_join(anno, by = c("sample" = "tube.id"))
cooks_tidy

# Summarize Cook's distance per sample
sample_cooks_summary <- cooks_tidy %>%
  group_by(sample, pop) %>%
  summarize(mean_cooks = mean(cooks_distance, na.rm = TRUE),
            pct95_cooks = quantile(cooks_distance, 0.99, na.rm = TRUE),
            .groups = "drop")
sample_cooks_summary %>%
  arrange(mean_cooks)
sample_cooks_summary
# Optionally: flag samples above a threshold
threshold <- 1  # can be adjusted

flagged <- sample_cooks_summary %>%
  filter(pct95_cooks > threshold) %>%
  arrange(desc(pct95_cooks))

# Output the flagged samples
print(flagged)

# Optional visualization
ggplot(sample_cooks_summary, aes(x = sample, y = pct95_cooks, fill = pop)) +
  geom_col() +
  geom_hline(yintercept = threshold, linetype = "dashed", color = "red") +
  labs(title = "95th Percentile of Cook's Distance by Sample",
       y = "95th Percentile Cook's Distance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



```{r deseq gill}
pcaData_gill %>%
  arrange(PC1)
pcaData_gill
anno

gill_PCA_for_sup <- ggplot(pcaData_gill, aes(x = PC1, y = PC2, color = habitat, shape =population)) + 
  geom_point(size =3, aes(fill=treatment)) + 
  geom_point(size =3) +
  scale_shape_manual(values = c(15,2, 0,17,)) +
  scale_color_manual(values=c('gold','skyblue')) +
  scale_fill_manual(values=gill_meta) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) + 
  ylab(paste0("PC2: ", percentVar[2], "% variance")) + 
  theme_minimal() +
  ggtitle("Gill PCA, n = 150") 

#pdf("/Users/kararyan/Desktop/Tagseq_exposure_may25/Gill_PCA.pdf")
gill_PCA
#pdf("/Users/kararyan/Desktop/Tagseq_exposure_may25/FigS2.pdf")
gill_PCA_for_sup

```

### set up for new contrasts

```{r deseq gill}
resultsNames(dds_gill)

mod_mat <- model.matrix(design(dds_gill), colData(dds_gill))

# calculate the vector of coefficient weights
# taco S
taco_s_full <- colMeans(mod_mat[dds_gill$group == 'taco.s.AP39.full',])
taco_s_con <- colMeans(mod_mat[dds_gill$group == 'taco.s.control',])
taco_s_pbs <- colMeans(mod_mat[dds_gill$group == 'taco.s.PBS',])

#taco NS
taco_ns_full <- colMeans(mod_mat[dds_gill$group == 'taco.ns.AP39.full',])
taco_ns_con <- colMeans(mod_mat[dds_gill$group == 'taco.ns.control',])
taco_ns_pbs <- colMeans(mod_mat[dds_gill$group == 'taco.ns.PBS',])

# puya S
puya_s_full <- colMeans(mod_mat[dds_gill$group == 'puya.s.AP39.full',])
puya_s_con <- colMeans(mod_mat[dds_gill$group == 'puya.s.control',])
puya_s_pbs <- colMeans(mod_mat[dds_gill$group == 'puya.s.PBS',])

# Puya NS
puya_ns_full <- colMeans(mod_mat[dds_gill$group == 'puya.ns.AP39.full',])
puya_ns_con <- colMeans(mod_mat[dds_gill$group == 'puya.ns.control',])
puya_ns_pbs <- colMeans(mod_mat[dds_gill$group == 'puya.ns.PBS',])
```

### Gill control vs PBS

```{r gill control pbs}
# Taco S (AvsB) 
Taco_s_A_B_gill <- results(dds_gill, contrast = taco_s_con - taco_s_pbs) 
sum(Taco_s_A_B_gill$padj < 0.05, na.rm=T)
Taco_s_A_B_gill <- as.data.frame(Taco_s_A_B_gill)
# = 9 

Taco_s_A_B_gill$gene <- rownames(Taco_s_A_B_gill)
Taco_s_A_B_gill %>%
  arrange(padj) %>% 
  mutate(diffexpressed =
                     case_when(log2FoldChange < 0 & padj < 0.05 ~ "Down", 
                               log2FoldChange > 0 & padj < 0.05 ~ "Up",
                               TRUE ~ "No")) %>% 
  left_join(Pmex_human,join_by('gene' =='gene.ID')) ->Taco_s_A_B_gill


# Taco NS (AvsB)
Taco_ns_A_B_gill <- results(dds_gill, contrast = taco_ns_con - taco_ns_pbs) 
sum(Taco_ns_A_B_gill$padj < 0.05, na.rm=T)
#= 19 
Taco_ns_A_B_gill <- as.data.frame(Taco_ns_A_B_gill)
Taco_ns_A_B_gill$gene <- rownames(Taco_ns_A_B_gill)
Taco_ns_A_B_gill %>%
  arrange(padj) %>% 
  left_join(Pmex_human,join_by('gene' =='gene.ID')) %>% 
  mutate(diffexpressed =
                     case_when(log2FoldChange < 0 & padj < 0.05 ~ "Down", 
                               log2FoldChange > 0 & padj < 0.05 ~ "Up",
                               TRUE ~ "No")) ->Taco_ns_A_B_gill
Taco_ns_A_B_gill

# puya S (AvsB)
puya_s_A_B_gill <- results(dds_gill, contrast = puya_s_con - puya_s_pbs) 
sum(puya_s_A_B_gill$padj < 0.05, na.rm=T)
# = 5 
puya_s_A_B_gill <- as.data.frame(puya_s_A_B_gill)
puya_s_A_B_gill %>%
  arrange(padj)

# puya NS (AvsB)
puya_ns_A_B_gill <- results(dds_gill, contrast = puya_ns_con - puya_ns_pbs) 
sum(puya_ns_A_B_gill$padj < 0.05, na.rm=T)

puya_ns_A_B_gill <- as.data.frame(puya_ns_A_B_gill)
puya_ns_A_B_gill %>%
  arrange(padj)

0+5+19+9
33/4
```



```{r}
# Taco NS vs S 
Taco_s_taco_ns_A <- results(dds_gill, contrast = ((taco_s_con + taco_s_pbs)/2) - ((taco_ns_con+ taco_ns_pbs)/2)) 
sum(Taco_s_taco_ns_A$padj < 0.05, na.rm=T)
Taco_s_taco_ns_A
# make the results a data frame
Taco_s_taco_ns_A  <- as.data.frame(Taco_s_taco_ns_A )
# make a column of the gene names 
Taco_s_taco_ns_A$gene <- rownames(Taco_s_taco_ns_A )
# join with annotations 
Taco_s_taco_ns_A  %>% 
  left_join(Pmex_human,join_by('gene' =='gene.ID')) ->Taco_s_taco_ns_A 
arrange(Taco_s_taco_ns_A, padj)
Taco_s_taco_ns_A  %>%
  arrange(padj)  %>% 
  mutate(diffexpressed =
                     case_when(log2FoldChange < 0 & padj < 0.05 ~ "Down", 
                               log2FoldChange > 0 & padj < 0.05 ~ "Up",
                               TRUE ~ "No"))->Taco_s_taco_ns_A 
Taco_s_taco_ns_A$diffexpressed <- as.factor(Taco_s_taco_ns_A$diffexpressed)
summary(Taco_s_taco_ns_A)

summary(Taco_s_taco_ns_A)

write.table(Taco_s_taco_ns_A , '/Users/kararyan/Documents/AP39_Tagseq/taco_A_ns_s_gill.tsv',quote = F, row.names = F,sep = '\t' )

Taco_s_ns_AP39 <- results(dds_gill, contrast = taco_s_full - taco_ns_full)
sum(Taco_s_ns_AP39$padj < 0.05, na.rm=T)
Taco_s_ns_AP39 <- as.data.frame(Taco_s_ns_AP39)
Taco_s_ns_AP39$gene <- rownames(Taco_s_ns_AP39 )
Taco_s_ns_AP39 %>% 
  left_join(Pmex_human,join_by('gene' =='gene.ID')) ->Taco_s_ns_AP39

Taco_s_ns_AP39%>%
  arrange(padj)  %>% 
  mutate(diffexpressed =
                     case_when(log2FoldChange < 0 & padj < 0.05 ~ "Down", 
                               log2FoldChange > 0 & padj < 0.05 ~ "Up",
                               TRUE ~ "No"))->Taco_s_ns_AP39
Taco_s_ns_AP39$diffexpressed <- as.factor(Taco_s_ns_AP39$diffexpressed)
summary(Taco_s_ns_AP39)

```



```{r}

# puya ecotype gill
puya_s_puya_ns_A<- results(dds_gill, contrast = ((puya_s_con+ puya_s_pbs)/2) - ((puya_ns_con+puya_ns_pbs)/2))
sum(puya_s_puya_ns_A$padj < 0.05, na.rm=T)
arrange(puya_s_puya_ns_A, padj)
sum(puya_s_puya_ns_A) %>%
      arrange(padj)
# make the results a data frame
puya_s_puya_ns_A  <- as.data.frame(puya_s_puya_ns_A )
# make a column of the gene names 
puya_s_puya_ns_A$gene <- rownames(puya_s_puya_ns_A )
# join with annotations 
puya_s_puya_ns_A  %>% 
  left_join(Pmex_human,join_by('gene' =='gene.ID')) ->puya_s_puya_ns_A 
puya_s_puya_ns_A
puya_s_puya_ns_A  %>%
  arrange(padj)  %>% 
  mutate(diffexpressed =
                     case_when(log2FoldChange < 0 & padj < 0.05 ~ "Down", 
                               log2FoldChange > 0 & padj < 0.05 ~ "Up",
                               TRUE ~ "No"))->puya_s_puya_ns_A 
puya_s_puya_ns_A$diffexpressed<- as.factor(puya_s_puya_ns_A$diffexpressed)
summary(puya_s_puya_ns_A)
puya_s_puya_ns_A  %>%
  filter(gene.name == 'LOC106917689')
puya_s_puya_ns_A
write.table(puya_s_puya_ns_A , '/Users/kararyan/Documents/AP39_Tagseq/puya_A_ns_s_gill.tsv',quote = F, row.names = F,sep = '\t' )

puya_s_puya_ns_A %>%
  filter(padj < 0.05) %>%
  inner_join(
    Taco_s_taco_ns_A %>% filter(padj < 0.05),
    by = join_by(gene, gene.name, Uniprot, Hugo_human, Protein.annotations)
  )
puya_s_puya_ns_A %>%
  filter(gene.name == "sqrdl" | gene.name == 'LOC106917689' | gene.name == 'LOC106917690')

puya_s_puya_ns_A %>%
  filter(padj < 0.05) -> puya_s_ns_sig_gill

Taco_s_taco_ns_A  %>%
  filter(padj < 0.05) -> taco_s_ns_sig_gill
taco_s_ns_sig_gill
puya_s_ns_sig_gill

puya_s_ns_sig_gill %>%
  inner_join(taco_s_ns_sig_gill, by = 'gene') %>%
  select(log2FoldChange.x, log2FoldChange.y, gene, gene.name.x, Hugo_human.x, Protein.annotations.x) -> gill_control_shared

capture_genes<- read.csv('/Users/kararyan/Downloads/mec17081-sup-0001-appendixs1/mec17081-sup-0002-TableS2.csv')
capture_genes
gill_control_shared
capture_genes %>%
  full_join(gill_control_shared, by = c('Gene.name.from.accession' = 'gene.name.x'))
```

```{r}
padj_cutoff <- 0.05
lfc_cutoff <- 0
up_taco_s_ns <- taco_s_ns_sig_gill  %>% filter(padj < padj_cutoff, log2FoldChange > lfc_cutoff) %>% pull(gene)
  
down_taco_s_ns <- taco_s_ns_sig_gill  %>% filter(padj < padj_cutoff, log2FoldChange < lfc_cutoff) %>% pull(gene)

up_puya_s_ns <- puya_s_puya_ns_A  %>% filter(padj < padj_cutoff, log2FoldChange > lfc_cutoff) %>% pull(gene)
  
down_puya_s_ns <- puya_s_puya_ns_A  %>% filter(padj < padj_cutoff, log2FoldChange < lfc_cutoff) %>% pull(gene)

# Create gene lists
up_list_eco <- list(
  Tacotalpa = up_taco_s_ns,
  Puyacatengo = up_puya_s_ns
)

down_list_eco <- list(
  Tacotalpa = down_taco_s_ns,
  Puyacatengo = down_puya_s_ns
)

venn_up_eco <- venn.diagram(
  x = up_list_eco,
  category.names = names(up_list_eco),
  filename = NULL,
 # fill = c("red", "orange", "tomato"),
  alpha = 0.5,
  cex = 1.2,
  fontface = "bold",
  cat.cex = 1.2,
  cat.fontface = "bold",
  main = "Upregulated Genes"
)
grid.newpage()
pdf(file="/Users/kararyan/Documents/AP39_Tagseq/gill_venn_up_eco.pdf")
grid.draw(venn_up_eco)
dev.off()

venn_down_eco <- venn.diagram(
  x = down_list_eco,
  category.names = names(down_list_eco),
  filename = NULL,
 # fill = c("red", "orange", "tomato"),
  alpha = 0.5,
  cex = 1.2,
  fontface = "bold",
  cat.cex = 1.2,
  cat.fontface = "bold",
  main = "Upregulated Genes"
)
grid.newpage()
pdf(file="/Users/kararyan/Documents/AP39_Tagseq/gill_venn_down_eco.pdf")
grid.draw(venn_down_eco)
dev.off()

library(gplots)
ItemsList <- venn(up_list_eco, show.plot = FALSE)

lengths(attributes(ItemsList)$intersections)
list_s_up<- attributes(ItemsList)$intersections$`Tacotalpa:Puyacatengo`
genes<- as.data.frame(list_s_up)
genes
taco_s_ns_sig_gill %>%
  inner_join(genes, by = c("gene" = "list_s_up")) ->eco_gill_2

ItemsList <- venn(down_list_eco, show.plot = FALSE)
lengths(attributes(ItemsList)$intersections)
list_s_down<- attributes(ItemsList)$intersections$`Tacotalpa:Puyacatengo`
genes<- as.data.frame(list_s_down)
genes
taco_s_ns_sig_gill %>%
  inner_join(genes, by = c("gene" = "list_s_down")) -> eco_gill_1

eco_gill_1 %>%
  merge(eco_gill_2)
rbind(eco_gill_2, eco_gill_1) -> eco_gill_genes_forground

eco_gill_genes_forground<- na.omit(eco_gill_genes_forground)
eco_gill_genes_forground

genes <- eco_gill_genes_forground$Unprot
genes
genes %>%
  dplyr::select(Uniprot)
# Name the vector
genes <- eco_gill_genes_forground$Uniprot
genes
# omit NA values
genes <- na.omit(genes)
genes
na.omit(puya_s_puya_ns_A)
na.omit(puya_s_puya_ns_A) ->puya_s_puya_ns_A.rm

library(org.Hs.eg.db, character.only = TRUE)

names(taco_s_ns_sig_gill)
go_enrich <- enrichGO(gene = genes,
                      universe = puya_s_puya_ns_A.rm$Uniprot,
                      OrgDb = org.Hs.eg.db::org.Hs.eg.db, 
                      keyType = 'UNIPROT',
                      readable = T,
                      ont = "ALL",
                      minGSSize = 2, 
                      pvalueCutoff = 0.05)
go_enrich 
dotplot(go_enrich)

?enrichGO
```





### Gill ecotype comparison (ap39 vs ap39)

```{r}
# Taco NS vs S 
Taco_s_taco_ns_D <- results(dds_gill, contrast = taco_s_full - taco_ns_full) 
sum(Taco_s_taco_ns_D$padj < 0.05, na.rm=T)

Taco_s_taco_ns_D
# make the results a data frame
Taco_s_taco_ns_D  <- as.data.frame(Taco_s_taco_ns_D )
# make a column of the gene names 
Taco_s_taco_ns_D$gene <- rownames(Taco_s_taco_ns_D )
# join with annotations 
Taco_s_taco_ns_D  %>% 
  left_join(Pmex_human,join_by('gene' =='gene.ID')) ->Taco_s_taco_ns_D 
arrange(Taco_s_taco_ns_D, padj)
Taco_s_taco_ns_D  %>%
  arrange(padj)   %>% 
  mutate(diffexpressed =
                     case_when(log2FoldChange < 0 & padj < 0.05 ~ "Down", 
                               log2FoldChange > 0 & padj < 0.05 ~ "Up",
                               TRUE ~ "No"))


Taco_s_taco_ns_D %>%
  filter(padj < 0.05) %>%
  inner_join(
    Taco_s_taco_ns_A %>% filter(padj < 0.05),
    by = join_by(gene, gene.name, Uniprot, Hugo_human, Protein.annotations)
  )
Taco_s_taco_ns_D 
write.table(Taco_s_taco_ns_D , '/Users/kararyan/Documents/AP39_Tagseq/taco_D_ns_s_gill.tsv',quote = F, row.names = F,sep = '\t' )
```

### Gill compare AP39 to control + PBS

```{r gill results}
# Taco S (A+B/2) vs D 
Taco_s_AB_D_gill <- results(dds_gill, contrast = taco_s_full - (taco_s_con + taco_s_pbs)/2) 
sum(Taco_s_AB_D_gill$padj < 0.05, na.rm=T)
Taco_s_AB_D_gill<- as.data.frame(Taco_s_AB_D_gill )
Taco_s_AB_D_gill$gene <- rownames(Taco_s_AB_D_gill)

Taco_s_AB_D_gill %>% 
  left_join(Pmex_human,join_by('gene' =='gene.ID')) %>%
  arrange(padj) %>%
  filter(padj < 0.05) -> Taco_s_AB_D_gill_sig


Taco_s_AB_D_gill
# Taco NS (A+B/2) vs D 
Taco_ns_AB_D_gill <- results(dds_gill, contrast =taco_ns_full - (taco_ns_con + taco_ns_pbs)/2) 
sum(Taco_ns_AB_D_gill$padj < 0.05, na.rm=T)
Taco_ns_AB_D_gill<- as.data.frame(Taco_ns_AB_D_gill)
Taco_ns_AB_D_gill$gene <- rownames(Taco_ns_AB_D_gill)
 Taco_ns_AB_D_gill%>%
  left_join(Pmex_human,join_by('gene' =='gene.ID')) %>%
  arrange(padj)  %>%
  filter(padj < 0.05) ->  Taco_ns_AB_D_gill_sig

# puya S (A+B/2) vs D
puya_s_AB_D_gill <- results(dds_gill, contrast = puya_s_full - (puya_s_con + puya_s_pbs)/2) 
sum(puya_s_AB_D_gill$padj < 0.05, na.rm=T)
puya_s_AB_D_gill<- as.data.frame(puya_s_AB_D_gill)
puya_s_AB_D_gill$gene <- rownames(puya_s_AB_D_gill)
puya_s_AB_D_gill %>%
  left_join(Pmex_human,join_by('gene' =='gene.ID')) %>%
  arrange(padj) %>%
  filter(padj < 0.05) -> puya_s_AB_D_gill_sig

# puya NS(A+B/2) vs D
puya_ns_AB_D_gill <- results(dds_gill, contrast =puya_ns_full - (puya_ns_con + puya_ns_pbs)/2) 
sum(puya_ns_AB_D_gill$padj < 0.05, na.rm=T)

puya_ns_AB_D_gill<- as.data.frame(puya_ns_AB_D_gill)
puya_ns_AB_D_gill$gene <- rownames(puya_ns_AB_D_gill)
 puya_ns_AB_D_gill %>%
   left_join(Pmex_human,join_by('gene' =='gene.ID')) %>%
  arrange(padj) %>%
  filter(padj < 0.05) -> puya_ns_AB_D_gill_sig
 
 
 puya_ns_AB_D_gill_sig
 puya_s_AB_D_gill_sig
 
 
puya_ns_AB_D_gill_sig %>%
inner_join(puya_s_AB_D_gill_sig, by = 'gene') %>%
  select(log2FoldChange.x, log2FoldChange.y, gene, gene.name.x, Hugo_human.x) -> shared_puya
shared_puya

Taco_ns_AB_D_gill_sig
Taco_s_AB_D_gill_sig

Taco_ns_AB_D_gill_sig%>%
  inner_join(Taco_s_AB_D_gill_sig, by = 'gene') %>%
  select(log2FoldChange.x, log2FoldChange.y, gene, gene.name.x, Hugo_human.x) -> shared_taco

shared_taco %>%
  inner_join(shared_puya, by = 'gene')
shared_puya
shared_taco

 puya_s_AB_D_gill_sig %>%
   inner_join(Taco_s_AB_D_gill_sig, by = 'gene')
 
 
  puya_ns_AB_D_gill_sig %>%
   inner_join(Taco_ns_AB_D_gill_sig, by = 'gene')
```

## make Venn of overlap
```{r gill results}
padj_cutoff <- 0.05
lfc_cutoff <- 0
up_puya_s <- puya_s_AB_D_gill_sig  %>% filter(padj < padj_cutoff, log2FoldChange > lfc_cutoff) %>% pull(gene)
  
down_puya_s <- puya_s_AB_D_gill_sig  %>% filter(padj < padj_cutoff, log2FoldChange < lfc_cutoff) %>% pull(gene)

up_puya_ns <- puya_ns_AB_D_gill_sig  %>% filter(padj < padj_cutoff, log2FoldChange > lfc_cutoff) %>% pull(gene)
  
down_puya_ns <- puya_ns_AB_D_gill_sig  %>% filter(padj < padj_cutoff, log2FoldChange < lfc_cutoff) %>% pull(gene)

up_taco_s <- Taco_s_AB_D_gill_sig  %>% filter(padj < padj_cutoff, log2FoldChange > lfc_cutoff) %>% pull(gene)
  
down_taco_s <- Taco_s_AB_D_gill_sig  %>% filter(padj < padj_cutoff, log2FoldChange < lfc_cutoff) %>% pull(gene)

up_taco_ns <- Taco_ns_AB_D_gill_sig  %>% filter(padj < padj_cutoff, log2FoldChange > lfc_cutoff) %>% pull(gene)
  
down_taco_ns <- Taco_ns_AB_D_gill_sig  %>% filter(padj < padj_cutoff, log2FoldChange < lfc_cutoff) %>% pull(gene)

# Create gene lists
up_list <- list(
  PuyaS = up_puya_s,
  TacoS = up_taco_s,
  PuyaNS = up_puya_ns,
  TacoNS = up_taco_ns
)

down_list <- list(
  PuyaS = down_puya_s,
  TacoS = down_taco_s,
  PuyaNS = down_puya_ns,
  TacoNS = down_taco_ns
)

down_list
# Generate Venn diagram for upregulated genes
venn_up <- venn.diagram(
  x = up_list,
  category.names = names(up_list),
  filename = NULL,
 # fill = c("red", "orange", "tomato"),
  alpha = 0.5,
  cex = 1.2,
  fontface = "bold",
  cat.cex = 1.2,
  cat.fontface = "bold",
  main = "Upregulated Genes"
)
grid.newpage()
pdf(file="/Users/kararyan/Documents/AP39_Tagseq/gill_venn_up.pdf")
grid.draw(venn_up)
dev.off()



library(gplots)
ItemsList <- venn(up_list, show.plot = FALSE)

lengths(attributes(ItemsList)$intersections)
attributes(ItemsList)$intersections
puya_s_AB_D_gill_sig %>%
  filter(gene == "gene22741")

gene8866
```


```{r gill results}
# Generate Venn diagram for downregulated genes
venn_down <- venn.diagram(
  x = down_list,
  category.names = names(down_list),
  filename = NULL,
 # fill = c("dodgerblue", "skyblue", "steelblue"),
  alpha = 0.5,
  cex = 1.2,
  fontface = "bold",
  cat.cex = 1.2,
  cat.fontface = "bold",
  main = "Downregulated Genes"
)
grid.newpage()
grid.newpage()
pdf(file="/Users/kararyan/Documents/AP39_Tagseq/gill_venn_down.pdf")
grid.draw(venn_down)
dev.off()

```

### 

### Annotate gill results and write tables

#### Taco

```{r annotate gill results}
# make the results a data frame
Taco_s_AB_D_gill <- as.data.frame(Taco_s_AB_D_gill)
# make a column of the gene names 
Taco_s_AB_D_gill$gene <- rownames(Taco_s_AB_D_gill)
# join with annotations 
Taco_s_AB_D_gill %>% 
  left_join(Pmex_human,join_by('gene' =='gene.ID')) ->Taco_s_AB_D_gill
Taco_s_AB_D_gill %>%
  arrange(padj)  %>% 
  mutate(diffexpressed =
                     case_when(log2FoldChange < 0 & padj < 0.05 ~ "Down", 
                               log2FoldChange > 0 & padj < 0.05 ~ "Up",
                               TRUE ~ "No"))->Taco_s_AB_D_gill

Taco_s_AB_D_gill
write.table(Taco_s_AB_D_gill, '/Users/kararyan/Documents/AP39_Tagseq/Taco_s_AB_D_gill_downsampled.tsv',quote = F, row.names = F,sep = '\t' )

# make the results a data frame
Taco_ns_AB_D_gill <- as.data.frame(Taco_ns_AB_D_gill)
# make a column of the gene names 
Taco_ns_AB_D_gill$gene <- rownames(Taco_ns_AB_D_gill)
# join with annotations 
Taco_ns_AB_D_gill %>% 
  left_join(Pmex_human,join_by('gene' =='gene.ID')) ->Taco_ns_AB_D_gill
Taco_ns_AB_D_gill %>%
  arrange(padj) -> Taco_ns_AB_D_gill

head(Taco_ns_AB_D_gill)
Taco_ns_AB_D_gill %>%
  arrange(padj)  %>% 
  mutate(diffexpressed =
                     case_when(log2FoldChange < 0 & padj < 0.05 ~ "Down", 
                               log2FoldChange > 0 & padj < 0.05 ~ "Up",
                               TRUE ~ "No"))->Taco_ns_AB_D_gill
Taco_ns_AB_D_gill
write.table(Taco_ns_AB_D_gill, '/Users/kararyan/Documents/AP39_Tagseq/Taco_ns_AB_D_gill_downsampled.tsv',quote = F, row.names = F,sep = '\t' )
```

#### Puya

```{r annotate gill results puya}
# make the results a data frame
puya_s_AB_D_gill <- as.data.frame(puya_s_AB_D_gill)
# make a column of the gene names 
puya_s_AB_D_gill$gene <- rownames(puya_s_AB_D_gill)
# join with annotations 
puya_s_AB_D_gill %>% 
  left_join(Pmex_human,join_by('gene' =='gene.ID')) ->puya_s_AB_D_gill
puya_s_AB_D_gill %>%
  arrange(padj)  %>% 
  mutate(diffexpressed =
                     case_when(log2FoldChange < 0 & padj < 0.05 ~ "Down", 
                               log2FoldChange > 0 & padj < 0.05 ~ "Up",
                               TRUE ~ "No"))->puya_s_AB_D_gill
puya_s_AB_D_gill
write.table(puya_s_AB_D_gill, '/Users/kararyan/Documents/AP39_Tagseq/puya_s_AB_D_gill_downsampled.tsv',quote = F, row.names = F,sep = '\t' )

# make the results a data frame
puya_ns_AB_D_gill <- as.data.frame(puya_ns_AB_D_gill)
# make a column of the gene names 
puya_ns_AB_D_gill$gene <- rownames(puya_ns_AB_D_gill)
# join with annotations 
puya_ns_AB_D_gill %>% 
  left_join(Pmex_human,join_by('gene' =='gene.ID')) ->puya_ns_AB_D_gill
puya_ns_AB_D_gill %>%
  arrange(padj) %>% 
  mutate(diffexpressed =
                     case_when(log2FoldChange < 0 & padj < 0.05 ~ "Down", 
                               log2FoldChange > 0 & padj < 0.05 ~ "Up",
                               TRUE ~ "No")) ->puya_ns_AB_D_gill
write.table(puya_ns_AB_D_gill, '/Users/kararyan/Documents/AP39_Tagseq/puya_ns_AB_D_gill_downsampled.tsv',quote = F, row.names = F,sep = '\t' )
```

## Liver comparisons

```{r gill set up deseq object liver}
# filter metadata to liver tissue only and remove sample swap
anno %>%
  filter(tissue =='liver' & tube.id != 'EFR22.081' & treatment != "AP39.half" & tube.id != 'EFR22.193') %>%
  select(tube.id, pop, treatment, group, mass, habitat, drainage)-> liver_meta

# create a list of tube.ids to keep 
keep <- liver_meta$tube.id

#filter counts matrix to keep samples in liver list
rownames(counts) <- counts$Geneid
counts %>%
  select(all_of(keep)) -> liver_counts
liver_counts <- liver_counts[rowSums(liver_counts > 5) > 7, ]

# get sample size of unfiltered liver data
ncol(liver_counts)

# convert liver counts to a matrix
liver_counts_matrix <- as.matrix(liver_counts)

# remove tube id column and convert it to the rownames 
liver_anno_matrix <- as.matrix(liver_meta[,-1])
rownames(liver_anno_matrix) <- liver_meta[,1]
liver_meta

# confirm the same number of samples in both the counts and annotation
all(rownames(liver_anno_matrix %in% colnames(liver_counts_matrix)))
# confirm the sample names are in the matrix 
all(rownames(liver_anno_matrix) == colnames(liver_counts_matrix))

liver_meta$pop <- as.factor(liver_meta$pop)
liver_meta$treatment <- as.factor(liver_meta$treatment)
liver_meta %>%
  group_by(pop) %>%
  group_split() %>%
  lapply(summary)
```

### Set up deseq dataset

```{r deseq liver}
# set up deseq2 dataset 
dds_liver <- DESeqDataSetFromMatrix(countData = liver_counts_matrix, colData = liver_anno_matrix, 
                              design = ~ mass + pop + treatment)
# Run DESeq2 
dds_liver <- DESeq(dds_liver)
colData(dds_liver)

vsd_liver <- vst(dds_liver, blind=T)

vst_cor <- cor(assay(vsd_liver))
average_cor <- as.data.frame(colMeans(vst_cor))

pcaData_liver<- plotPCA(vsd_liver, intgroup=c("treatment", 'pop'), returnData=TRUE, ntop = 10000) 
percentVar <- round(100 * attr(pcaData_liver, "percentVar")) 
pcaData_liver$fill_color <- ifelse(pcaData_liver$treatment == "AP39.full",
                                  pcaData_liver$habitat,
                                  NA)

# Step 2: Define color and fill palettes
color_palette <- c("s" = "gold", "ns" = "skyblue")  # change to actual habitat names
shape_palette <- c("taco" = 22, "puya" = 24)
pcaData_liver
# Use shape values 21 and 24 (can be filled)
liver_PCA <- ggplot(pcaData_liver, aes(x = PC1, y = PC2,
                                     color = habitat,
                                     shape = drainage,
                                     fill = fill_color)) +
  geom_point(size = 3, stroke = 1) +
#  geom_text(aes(label = name)) +
  scale_shape_manual(values = shape_palette) +
  scale_color_manual(values = color_palette) +
  scale_fill_manual(values = color_palette, na.value = NA) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme_minimal() +
  ggtitle("Liver")
liver_PCA 

pcaData_liver %>%
  arrange(PC1)
liver_PCA

pcaData_liver

```
## heatmap 
```{r}

dds_liver <- estimateSizeFactors(dds_liver)
# Variance Stabilizing Transformation
vsd_liver <- vst(dds_liver, blind = TRUE)

# Extract the transformed expression matrix
vsd_mat_liver <- assay(vsd_liver)

# Calculate the row variance for each gene
gene_variances_liver <- apply(vsd_mat_liver, 1, var)

# Select top 2500 most variable genes
top_genes_liver <- head(order(gene_variances_liver, decreasing = TRUE), 2000)
top_vsd_mat_liver <- vsd_mat_liver[top_genes_liver, ]

# Optional: Add annotation for columns (samples)
# Here we use the colData from the DESeq object
sample_anno_liver <- as.data.frame(colData(dds_liver))

# Generate the heatmap with hierarchical clustering
sample_anno_liver %>%
  select(treatment, habitat, drainage) -> sample_anno_2_liver

#pdf("heatmap_top2000.pdf", width = 10, height = 12)
pheatmap(top_vsd_mat_liver,
         scale = "row",                 # Standardize each gene (z-score)
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "complete",
         annotation_col = sample_anno_2_liver, # Sample metadata
         show_rownames = FALSE,        # Hide gene names for clarity
         fontsize_col = 10,            # Sample name font size
         main = "Top 2000 Variable Genes Heatmap")

# Save as PDF if desired

# pheatmap(...)  # Repeat the call inside
#dev.off()
```






### Design new contrasts

```{r deseq liver}
resultsNames(dds_liver)

mod_mat <- model.matrix(design(dds_liver), colData(dds_liver))

# calculate the vector of coefficient weights
taco_s_full <- colMeans(mod_mat[dds_liver$group == 'taco.s.AP39.full',])
taco_s_con <- colMeans(mod_mat[dds_liver$group == 'taco.s.control',])
taco_s_pbs <- colMeans(mod_mat[dds_liver$group == 'taco.s.PBS',])

taco_ns_full <- colMeans(mod_mat[dds_liver$group == 'taco.ns.AP39.full',])
taco_ns_con <- colMeans(mod_mat[dds_liver$group == 'taco.ns.control',])
taco_ns_pbs <- colMeans(mod_mat[dds_liver$group == 'taco.ns.PBS',])

puya_s_full <- colMeans(mod_mat[dds_liver$group == 'puya.s.AP39.full',])
puya_s_con <- colMeans(mod_mat[dds_liver$group == 'puya.s.control',])
puya_s_pbs <- colMeans(mod_mat[dds_liver$group == 'puya.s.PBS',])

puya_ns_full <- colMeans(mod_mat[dds_liver$group == 'puya.ns.AP39.full',])
puya_ns_con <- colMeans(mod_mat[dds_liver$group == 'puya.ns.control',])
puya_ns_pbs <- colMeans(mod_mat[dds_liver$group == 'puya.ns.PBS',])
```

### obtain results of new contrasts liver

```{r liver results}
# Taco S (A+B/2) vs D 
Taco_s_AB_D_liver <- results(dds_liver, alpha=0.1, contrast = taco_s_full - (taco_s_con + taco_s_pbs)/2) 
sum(Taco_s_AB_D_liver$padj < 0.05, na.rm=T)

# Taco NS (A+B/2) vs D 
Taco_ns_AB_D_liver <- results(dds_liver, alpha=0.1, contrast =taco_ns_full - (taco_ns_con + taco_ns_pbs)/2) 
sum(Taco_ns_AB_D_liver$padj < 0.05, na.rm=T)
sum(is.na(Taco_ns_AB_D_liver$padj))
Taco_ns_AB_D_liver

# puya S (A+B/2) vs D
puya_s_AB_D_liver <- results(dds_liver, alpha=0.1, contrast = puya_s_full - (puya_s_con + puya_s_pbs)/2) 
sum(puya_s_AB_D_liver$padj < 0.05, na.rm=T)

# puya NS(A+B/2) vs D
puya_ns_AB_D_liver <- results(dds_liver, alpha=0.1, contrast =puya_ns_full - (puya_ns_con + puya_ns_pbs)/2)
sum(puya_ns_AB_D_liver$padj < 0.05, na.rm=T)

puya_ns_s_con_liver <- results(dds_liver, contrast =((puya_ns_con + puya_ns_pbs)/2) - ((puya_s_con + puya_s_pbs)/2))
sum(puya_ns_s_con_liver$padj < 0.05, na.rm=T)


```




### annotate liver results and write tables

#### Taco

```{r annotate liver results}
# make the results a data frame
Taco_s_AB_D_liver <- as.data.frame(Taco_s_AB_D_liver)
# make a column of the gene names 
Taco_s_AB_D_liver$gene <- rownames(Taco_s_AB_D_liver)
# join with annotations 
Taco_s_AB_D_liver %>% 
  left_join(Pmex_human,join_by('gene' =='gene.ID')) ->Taco_s_AB_D_liver
Taco_s_AB_D_liver %>%
  arrange(desc(padj))  %>% 
  mutate(diffexpressed =
                     case_when(log2FoldChange < 0 & padj < 0.05 ~ "Down", 
                               log2FoldChange > 0 & padj < 0.05 ~ "Up",
                               TRUE ~ "No"))->Taco_s_AB_D_liver
write.table(Taco_s_AB_D_liver, '/Users/kararyan/Documents/AP39_Tagseq/Taco_s_AB_D_liver_downsampled.tsv',quote = F, row.names = F,sep = '\t' )

Taco_s_AB_D_liver
Taco_s_AB_D_liver %>%
  filter(gene.name == 'LOC106917690')
# make the results a data frame
Taco_ns_AB_D_liver <- as.data.frame(Taco_ns_AB_D_liver)
# make a column of the gene names 
Taco_ns_AB_D_liver$gene <- rownames(Taco_ns_AB_D_liver)
# join with annotations 
Taco_ns_AB_D_liver %>% 
  left_join(Pmex_human,join_by('gene' =='gene.ID')) ->Taco_ns_AB_D_liver
Taco_ns_AB_D_liver %>%
  arrange(desc(padj))  %>% 
  mutate(diffexpressed =
                     case_when(log2FoldChange < 0 & padj < 0.05 ~ "Down", 
                               log2FoldChange > 0 & padj < 0.05 ~ "Up",
                               TRUE ~ "No"))->Taco_ns_AB_D_liver

Taco_ns_AB_D_liver %>%
  filter(gene.name == 'sqrdl')
write.table(Taco_ns_AB_D_liver, '/Users/kararyan/Documents/AP39_Tagseq/Taco_ns_AB_D_liver_downsampled.tsv',quote = F, row.names = F,sep = '\t' )

```


```{r}

puya_ns_s_con_liver %>%
  arrange(padj)

puya_ns_s_con_liver<- as.data.frame(puya_ns_s_con_liver)
# make a column of the gene names 
puya_ns_s_con_liver$gene <- rownames(puya_ns_s_con_liver)
puya_ns_s_con_liver
# join with annotations 
puya_ns_s_con_liver %>% 
  left_join(Pmex_human,join_by('gene' =='gene.ID')) ->puya_ns_s_con_liver 
puya_ns_s_con_liver  %>%
  arrange(desc(padj))  %>% 
  mutate(diffexpressed =
                     case_when(log2FoldChange < 0 & padj < 0.05 ~ "Down", 
                               log2FoldChange > 0 & padj < 0.05 ~ "Up",
                               TRUE ~ "No"))->puya_ns_s_con_liver 
puya_ns_s_con_liver

puya_ns_s_con_liver  %>%
  filter(gene.name == 'LOC106917690')
```

#### Puya

```{r annotate liver results puya}

# make the results a data frame
puya_s_AB_D_liver <- as.data.frame(puya_s_AB_D_liver)
# make a column of the gene names 
puya_s_AB_D_liver$gene <- rownames(puya_s_AB_D_liver)
# join with annotations 
puya_s_AB_D_liver %>% 
  left_join(Pmex_human,join_by('gene' =='gene.ID')) ->puya_s_AB_D_liver
puya_s_AB_D_liver %>%
  arrange(desc(padj)) %>% 
  mutate(diffexpressed =
                     case_when(log2FoldChange < 0 & padj < 0.05 ~ "Down", 
                               log2FoldChange > 0 & padj < 0.05 ~ "Up",
                               TRUE ~ "No")) ->puya_s_AB_D_liver
write.table(puya_s_AB_D_liver, '/Users/kararyan/Documents/AP39_Tagseq/puya_s_AB_D_liver_downsampled.tsv',quote = F, row.names = F,sep = '\t' )
puya_s_AB_D_liver
puya_s_AB_D_liver %>%
  filter(gene.name == 'sqrdl')
# make the results a data frame
puya_ns_AB_D_liver <- as.data.frame(puya_ns_AB_D_liver)
# make a column of the gene names 
puya_ns_AB_D_liver$gene <- rownames(puya_ns_AB_D_liver)
# join with annotations 
puya_ns_AB_D_liver %>% 
  left_join(Pmex_human,join_by('gene' =='gene.ID')) ->puya_ns_AB_D_liver
puya_ns_AB_D_liver %>%
  arrange(desc(padj))  %>% 
  mutate(diffexpressed =
                     case_when(log2FoldChange < 0 & padj < 0.05 ~ "Down", 
                               log2FoldChange > 0 & padj < 0.05 ~ "Up",
                               TRUE ~ "No"))->puya_ns_AB_D_liver
write.table(puya_ns_AB_D_liver, '/Users/kararyan/Documents/AP39_Tagseq/puya_ns_AB_D_liver_downsampled.tsv',quote = F, row.names = F,sep = '\t' )

puya_ns_AB_D_liver %>%
  filter(gene.name == 'sqrdl')
```

```{r gill results}
padj_cutoff <- 0.05
lfc_cutoff <- 0

up_puya_s <- puya_s_AB_D_liver  %>% filter(padj < padj_cutoff, log2FoldChange > lfc_cutoff) %>% pull(gene)
  
down_puya_s <- puya_s_AB_D_liver  %>% filter(padj < padj_cutoff, log2FoldChange < lfc_cutoff) %>% pull(gene)

up_puya_ns <- puya_ns_AB_D_liver  %>% filter(padj < padj_cutoff, log2FoldChange > lfc_cutoff) %>% pull(gene)
  
down_puya_ns <- puya_ns_AB_D_liver  %>% filter(padj < padj_cutoff, log2FoldChange < lfc_cutoff) %>% pull(gene)

up_taco_s <- Taco_s_AB_D_liver  %>% filter(padj < padj_cutoff, log2FoldChange > lfc_cutoff) %>% pull(gene)
  
down_taco_s <- Taco_s_AB_D_liver %>% filter(padj < padj_cutoff, log2FoldChange < lfc_cutoff) %>% pull(gene)

up_taco_ns <- Taco_ns_AB_D_liver  %>% filter(padj < padj_cutoff, log2FoldChange > lfc_cutoff) %>% pull(gene)
  
down_taco_ns <- Taco_ns_AB_D_liver  %>% filter(padj < padj_cutoff, log2FoldChange < lfc_cutoff) %>% pull(gene)
# Create gene lists
up_list <- list(
  PuyaS = up_puya_s,
  TacoS = up_taco_s,
  PuyaNS = up_puya_ns,
  TacoNS = up_taco_ns
)

down_list <- list(
  PuyaS = down_puya_s,
  TacoS = down_taco_s,
  PuyaNS = down_puya_ns,
  TacoNS = down_taco_ns
)

down_list
# Generate Venn diagram for upregulated genes
venn_up <- venn.diagram(
  x = up_list,
  category.names = names(up_list),
  filename = NULL,
 # fill = c("red", "orange", "tomato"),
  alpha = 0.5,
  cex = 1.2,
  fontface = "bold",
  cat.cex = 1.2,
  cat.fontface = "bold",
  main = "Upregulated Genes"
)
grid.newpage()
grid.draw(venn_up)

```


```{r gill results}
# Generate Venn diagram for downregulated genes
venn_down <- venn.diagram(
  x = down_list,
  category.names = names(down_list),
  filename = NULL,
 # fill = c("dodgerblue", "skyblue", "steelblue"),
  alpha = 0.5,
  cex = 1.2,
  fontface = "bold",
  cat.cex = 1.2,
  cat.fontface = "bold",
  main = "Downregulated Genes"
)
grid.newpage()
venn_down_liver<- grid.draw(venn_down)
venn_down_liver
```



## Enrichment 
```{r}
ItemsList <- venn(down_list, show.plot = FALSE)
Taco_ns_AB_D_liver
lengths(attributes(ItemsList)$intersections)
list_down<- attributes(ItemsList)$intersections$`PuyaS:TacoS:PuyaNS:TacoNS`
genes<- as.data.frame(list_down)
Taco_ns_AB_D_liver %>%
  inner_join(genes, by = c("gene" = "list_down")) -> liver_down

ItemsList <- venn(up_list, show.plot = FALSE)
lengths(attributes(ItemsList)$intersections)
list_up<- attributes(ItemsList)$intersections$`PuyaS:TacoS:PuyaNS:TacoNS`
genes<- as.data.frame(list_up)
Taco_ns_AB_D_liver %>%
  inner_join(genes, by = c("gene" = "list_up")) -> liver_up

rbind(liver_up, liver_down) -> liver_genes_forground

liver_genes_forground<- na.omit(liver_genes_forground)
liver_genes_forground

genes <- liver_genes_forground$Unprot

genes %>%
  dplyr::select(Uniprot)
# Name the vector
genes <- liver_genes_forground$Uniprot

# omit NA values
genes <- na.omit(genes)

library(org.Hs.eg.db, character.only = TRUE)

go_enrich <- enrichGO(gene = genes,
                      universe = Taco_ns_AB_D_liverA$Uniprot,
                      OrgDb = org.Hs.eg.db::org.Hs.eg.db, 
                      keyType = 'UNIPROT',
                      readable = T,
                      ont = "ALL",
                      pAdjustMethod = "fdr",
                      minGSSize = 2, 
                      pvalueCutoff = 0.05)

go_enrich
go_enrich$Description 
pdf('/Users/kararyan/Documents/AP39_Tagseq/shared_liver_ancestral.pdf', width = 6, height = 8)
dotplot(go_enrich, showCategory=20)
cnetplot(go_enrich)
```


```{r}

ItemsList <- venn(down_list, show.plot = FALSE)

lengths(attributes(ItemsList)$intersections)
list_down_s<- attributes(ItemsList)$intersections$`PuyaS:TacoS`
genes_s<- as.data.frame(list_down_s)
Taco_ns_AB_D_liver %>%
  inner_join(genes_s, by = c("gene" = "list_down_s")) -> liver_down_s

ItemsList <- venn(up_list, show.plot = FALSE)
lengths(attributes(ItemsList)$intersections)
list_up_s<- attributes(ItemsList)$intersections$`PuyaS:TacoS`
genes_s<- as.data.frame(list_up_s)
Taco_ns_AB_D_liver %>%
  inner_join(genes_s, by = c("gene" = "list_up_s")) -> liver_up_s

ItemsList <- venn(down_list, show.plot = FALSE)
lengths(attributes(ItemsList)$intersections)
list_down_ns<- attributes(ItemsList)$intersections$`PuyaNS:TacoNS`
genes_ns<- as.data.frame(list_down_ns)
Taco_ns_AB_D_liver %>%
  inner_join(genes_ns, by = c("gene" = "list_down_ns")) -> liver_down_ns

ItemsList <- venn(up_list, show.plot = FALSE)
lengths(attributes(ItemsList)$intersections)
list_up_ns<- attributes(ItemsList)$intersections$`PuyaNS:TacoNS`
genes_ns<- as.data.frame(list_up_ns)
Taco_ns_AB_D_liver %>%
  inner_join(genes_ns, by = c("gene" = "list_up_ns")) -> liver_up_ns

liver_list <- list(
  down_ancestral = liver_down,
  up_ancestral = liver_up,
  down_sulfide = liver_down_s$Uniprot,
  up_sulfide = liver_up_s$Uniprot,
  down_nonsulfide = liver_down_ns$Uniprot,
  up_nonsulfide = liver_up_ns$Uniprot
)
liver_down
liver_df <- bind_rows(
  tibble(UNIPROT = liver_down$Uniprot, group = "ancestral", direction = "down"),
  tibble(UNIPROT = liver_up$Uniprot, group = "ancestral", direction = "up"),
  tibble(UNIPROT = liver_down_s$Uniprot, group = "sulfide", direction = "down"),
  tibble(UNIPROT = liver_up_s$Uniprot, group = "sulfide", direction = "up"),
  tibble(UNIPROT = liver_down_ns$Uniprot, group = "nonsulfide", direction = "down"),
  tibble(UNIPROT = liver_up_ns$Uniprot, group = "nonsulfide", direction = "up")
)
formula_res <- compareCluster(UNIPROT~group+direction, data=liver_df, fun="enrichGO")
liver_df

formula_res <- compareCluster(
  UNIPROT ~ group + direction,
  data = liver_df,
  universe = Taco_ns_AB_D_liver$Uniprot,
  fun = "enrichGO",
  OrgDb = org.Hs.eg.db::org.Hs.eg.db,
  keyType = "UNIPROT", 
  pAdjustMethod = "fdr",
  ont="MF",
  minGSSize = 2,
  pvalueCutoff = 0.05,
  qvalueCutoff = 0.05, pool = F
)

pdf('/Users/kararyan/Documents/AP39_Tagseq/shared_liver_GO.pdf', width = 10, height = 10)
dotplot(formula_res)

liver_df_up <- bind_rows(
  tibble(UNIPROT = liver_up$Uniprot, group = "ancestral", direction = "up"),
  tibble(UNIPROT = liver_up_s$Uniprot, group = "sulfide", direction = "up"),
  tibble(UNIPROT = liver_up_ns$Uniprot, group = "nonsulfide", direction = "up")
)

liver_df_down <- bind_rows(
  tibble(UNIPROT = liver_down$Uniprot, group = "ancestral", direction = "down"),
  tibble(UNIPROT = liver_down_s$Uniprot, group = "sulfide", direction = "down"),
  tibble(UNIPROT = liver_down_ns$Uniprot, group = "nonsulfide", direction = "down"),
)

formula_res_up <- compareCluster(
  UNIPROT ~ group,
  data = liver_df_up,
  universe = puya_s_AB_D_liver$Uniprot,
  fun = "enrichGO",
  OrgDb = org.Hs.eg.db::org.Hs.eg.db,
  keyType = "UNIPROT", 
  pAdjustMethod = "fdr",
  ont="ALL",
  minGSSize = 5,
  pvalueCutoff = 0.05,
  pool = F
)

dotplot(formula_res_up)
head(formula_res_up, n = 500)
```


```{r}
Taco_s_AB_D_liver %>%
  filter(diffexpressed == "Up") -> Taco_s_AB_D_liver_up
Taco_s_AB_D_liver %>%
  filter(diffexpressed == "Down") -> Taco_s_AB_D_liver_down

Taco_ns_AB_D_liver %>%
  filter(diffexpressed == "Up") -> Taco_ns_AB_D_liver_up
Taco_ns_AB_D_liver %>%
  filter(diffexpressed == "Down") -> Taco_ns_AB_D_liver_down

puya_s_AB_D_liver %>%
  filter(diffexpressed == "Up") -> puya_s_AB_D_liver_up
puya_s_AB_D_liver %>%
  filter(diffexpressed == "Down") -> puya_s_AB_D_liver_down

puya_ns_AB_D_liver %>%
  filter(diffexpressed == "Up") -> puya_ns_AB_D_liver_up
puya_ns_AB_D_liver %>%
  filter(diffexpressed == "Down") -> puya_ns_AB_D_liver_down

liver_df <- bind_rows(
  tibble(UNIPROT = Taco_s_AB_D_liver_up$Uniprot, group = "taco_s", direction = "up"),
  tibble(UNIPROT = Taco_s_AB_D_liver_down$Uniprot, group = "taco_s", direction = "down"),
  tibble(UNIPROT = Taco_ns_AB_D_liver_up$Uniprot, group = "taco_ns", direction = "up"),
  tibble(UNIPROT = Taco_ns_AB_D_liver_down$Uniprot, group = "taco_ns", direction = "down"),
  tibble(UNIPROT = puya_s_AB_D_liver_up$Uniprot, group = "puya_s", direction = "up"),
  tibble(UNIPROT = puya_s_AB_D_liver_down$Uniprot, group = "puya_s", direction = "down"),
  tibble(UNIPROT = puya_ns_AB_D_liver_up$Uniprot, group = "puya_ns", direction = "up"),
  tibble(UNIPROT = puya_ns_AB_D_liver_down$Uniprot, group = "puya_ns", direction = "down"),
)

formula_res <- compareCluster(
  UNIPROT ~ group + direction,
  data = liver_df,
  universe = Taco_ns_AB_D_liver$Uniprot,
  fun = "enrichGO",
  OrgDb = org.Hs.eg.db::org.Hs.eg.db,
  keyType = "UNIPROT", 
  pAdjustMethod = "BH",
  ont="BP",
  minGSSize = 5,
  pvalueCutoff = 0.0001,
  qvalueCutoff = 0.0001, pool = F
)
dotplot(formula_res)

```






```{r}

Taco_s_AB_D_liver_rmna <- na.omit(Taco_s_AB_D_liver)
Taco_s_AB_D_liver_rmna <- unique(Taco_s_AB_D_liver_rmna$Uniprot)

unique(Taco_s_AB_D_liver_rmna)

taco_s_liver_ap39_enrich <- Taco_s_AB_D_liver_rmna$log2FoldChange 
Taco_s_AB_D_liver_rmna
names(taco_s_liver_ap39_enrich) <- Taco_s_AB_D_liver_rmna$Uniprot

taco_s_liver_ap39_enrich <- na.omit(taco_s_liver_ap39_enrich)

Taco_s_AB_D_liver_DE <- sort(Taco_s_AB_D_liver_DE, decreasing = TRUE)
deg_results_list_taco_s


taco_s_liver_ap39_enrich
gse_taco_s <- gseGO(gene = taco_s_liver_ap39_enrich,
             ont ="ALL", 
             keyType = "UNIPROT", 
             minGSSize = 3, 
             maxGSSize = 800, 
             pvalueCutoff = 0.05, 
             verbose = TRUE, 
             OrgDb = org.Hs.eg.db, 
             pAdjustMethod = "BH")


deg_results_list_taco_s

enrichGO()

require(DOSE)
dotplot(gse_taco_s, showCategory=10, split=".sign") + facet_grid(.~.sign)
```













# Liver PBS vs Control

```{r liver results}
# Taco S A vs C 
Taco_s_A_B_liver <- results(dds_liver, contrast = taco_s_pbs - taco_s_con) 
sum(Taco_s_A_B_liver$padj < 0.05, na.rm=T)

# Taco NS A vs C 
Taco_ns_A_B_liver <- results(dds_liver, contrast = taco_ns_pbs - taco_ns_con ) 
sum(Taco_ns_A_B_liver$padj < 0.05, na.rm=T)

# puya S A vs C 
puya_s_A_B_liver <- results(dds_liver, contrast = puya_s_pbs - puya_s_con ) 
sum(puya_s_A_B_liver$padj < 0.05, na.rm=T)

# puya NS A vs C 
puya_ns_A_B_liver <- results(dds_liver, contrast = puya_ns_pbs - puya_ns_con ) 
sum(puya_ns_A_B_liver$padj < 0.05, na.rm=T)


101 + 40 + 3 + 76 

220 / 4
```



```{r ecotype liver}
# Taco NS vs Scontrol
Taco_s_taco_ns_AB <- results(dds_liver, contrast = ((taco_s_con+ taco_s_pbs)/2) - ((taco_ns_con+ taco_ns_pbs)/2)) 
sum(Taco_s_taco_ns_AB$padj < 0.05, na.rm=T)

# make the results a data frame
Taco_s_taco_ns_AB  <- as.data.frame(Taco_s_taco_ns_AB )
# make a column of the gene names 
Taco_s_taco_ns_AB$gene <- rownames(Taco_s_taco_ns_AB )
# join with annotations 
Taco_s_taco_ns_AB  %>% 
  left_join(Pmex_human,join_by('gene' =='gene.ID')) ->Taco_s_taco_ns_AB 
Taco_s_taco_ns_AB  %>%
  arrange(padj)  %>% 
  mutate(diffexpressed =
                     case_when(log2FoldChange < 0 & padj < 0.05 ~ "Down", 
                               log2FoldChange > 0 & padj < 0.05 ~ "Up",
                               TRUE ~ "No"))->Taco_s_taco_ns_AB 
Taco_s_taco_ns_AB$diffexpressed <- as.factor(Taco_s_taco_ns_AB$diffexpressed) 
summary(Taco_s_taco_ns_AB)
#write.table(Taco_s_taco_ns_AB, '/Users/kararyan/Desktop/Tagseq_exposure/tables2/taco_A_ns_s.tsv',quote = F, row.names = F,sep = '\t' )
```


```{r ecotype liver}
# puya ecotype liverr
puya_s_puya_ns_AB<- results(dds_liver, contrast = ((puya_s_con+ puya_s_pbs)/2) - ((puya_ns_con + puya_ns_pbs)/2)) 
sum(puya_s_puya_ns_AB$padj < 0.05, na.rm=T)


# make the results a data frame
puya_s_puya_ns_AB  <- as.data.frame(puya_s_puya_ns_AB )
# make a column of the gene names 
puya_s_puya_ns_AB$gene <- rownames(puya_s_puya_ns_AB )
# join with annotations 
puya_s_puya_ns_AB  %>% 
  left_join(Pmex_human,join_by('gene' =='gene.ID')) ->puya_s_puya_ns_AB 
puya_s_puya_ns_AB  %>%
  arrange(desc(padj)) ->puya_s_puya_ns_AB 


puya_s_puya_ns_AB  %>% 
  mutate(diffexpressed =
                     case_when(log2FoldChange < 0 & padj < 0.05 ~ "Down", 
                               log2FoldChange > 0 & padj < 0.05 ~ "Up",
                               TRUE ~ "No")) -> puya_s_puya_ns_AB
 puya_s_puya_ns_AB$diffexpressed <- as.factor(puya_s_puya_ns_AB$diffexpressed)
summary(puya_s_puya_ns_AB)
# write.table(puya_s_puya_ns_AB , '/Users/kararyan/Desktop/Tagseq_exposure/tables2/puya_A_ns_s.tsv',quote = F, row.names = F,sep = '\t' )
puya_s_puya_ns_AB 

puya_s_puya_ns_AB %>%
  filter(padj < 0.05) -> puya_s_puya_ns_AB_liver_sig
Taco_s_taco_ns_AB %>%
  filter(padj < 0.05) -> Taco_s_taco_ns_AB_liver_sig

puya_s_puya_ns_AB_liver_sig %>%
  inner_join(Taco_s_taco_ns_AB_liver_sig, by = 'gene') %>%
  select(log2FoldChange.x, log2FoldChange.y, gene, Hugo_human.x,Protein.annotations.x )
```

Make PCAs for figure 1

```{r}
anno
anno %>%
  filter(tissue =='gill' & tube.id != 'EFR22.080' & treatment != "AP39.half") %>%
  select(tube.id, pop, treatment, group, mass, drainage)-> gill_meta

# Set up DESeq2 dataset
dds_gill <- DESeqDataSetFromMatrix(countData = gill_counts_matrix,
                              colData = gill_anno_matrix,
                              design = ~ mass + pop + treatment)

# Run DESeq2
dds_gill <- DESeq(dds_gill)

gill_anno<- as.data.frame(gill_anno_matrix)
# Subset samples (change condition as needed)
gill_anno_matrix
subset_condition <- gill_anno$drainage == "puya"  # example condition
dds_subset <- dds_gill[, subset_condition]
dds_subset
# Variance stabilizing transformation on subset
vsd_subset <- vst(dds_subset)
  
# PCA plot data for subset
pcaData_subset <- plotPCA(vsd_subset, intgroup = c("treatment", "pop"), returnData = TRUE, ntop = 100)
percentVar <- round(100 * attr(pcaData_subset, "percentVar"))

# Plot PCA
pca_plot_gill_puya <- ggplot(pcaData_subset, aes(x = PC1, y = PC2, color = treatment, shape = pop)) +
  geom_point(size = 3, aes(fill = treatment)) +
  scale_shape_manual(values = c(0, 17, 15, 2)) +
  scale_color_manual(values = c('gold', 'skyblue', 'skyblue')) +
  scale_fill_manual(values = gill_meta) +  # Ensure `gill_meta` still applies to the subset
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  theme_minimal() +
  ggtitle("Gill PCA (Subset), n = 150")

# Print plot
print(pca_plot_gill_taco)
print(pca_plot_gill_puya)


```

```{r}
library(ggpubr)
fig1_pca<- ggarrange(gill_PCA, liver_PCA, ncol=2,  common.legend = TRUE, legend="top")
pdf("/Users/kararyan/Documents/AP39_Tagseq/Figure1.pdf", height = 5, width = 9)
fig1_pca

```

